---
title: "STARsolo Cell Count Analysis - patient_3_ccRCC2_SRR12066652"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(Matrix)
library(ggplot2)
library(scales)
library(DropletUtils)
library(edgeR)
```

In this tutorial, we will demonstrate how to perform cell count analysis as being used in STARsolo,
the algorithm that STARsolo uses in order to make the cell filter (using EmptyDrops_CR for soloCellFilter parameter) is described in the following link:
https://genomebiology.biomedcentral.com/articles/10.1186/s13059-019-1662-y

# Load and Process STARsolo Data
```{r}
# Load the dataset
ccRCC.data <- Read10X(data.dir = "./raw")
# Initialize the Seurat object with the raw (non-normalized data).
ccRCC_p3 <- CreateSeuratObject(counts = ccRCC.data, project = "ccRCC_p3")
ccRCC_p3
```

# basic Analysis and Visualization

Step 1: Calculate total UMI counts per barcode

```{r}

# Ensure we are working with the RNA assay
DefaultAssay(ccRCC_p3) <- "RNA"

# Now calculate the total UMI counts
# Using the 'layer' argument as per the updated Seurat syntax
total_umi_counts <- Matrix::colSums(GetAssayData(ccRCC_p3, layer = "counts"))
total_umi_counts[1:10]

```

Step 2: Sort barcodes by total UMI counts
```{r}
sorted_umi_counts <- sort(total_umi_counts, decreasing = TRUE)
sorted_umi_counts[1:10]


# Find the index of the first barcode with a zero count
zero_count_index <- which(sorted_umi_counts == 0)[1]

```

Step 3: Visualizing the results up to the first zero count


```{r}
# If there is no zero count, plot all
if (!is.na(zero_count_index)) {
    # Filter the counts to include only up to the first zero count
    plot_umi_counts <- sorted_umi_counts[1:(zero_count_index-1)]
} else {
    plot_umi_counts <- sorted_umi_counts
}
```

```{r}

# Assuming 'sorted_umi_counts' contains your UMI counts per barcode

# Create a dataframe for plotting
barcode_data <- data.frame(
  Barcode = seq_along(sorted_umi_counts),
  UMI_Counts = sorted_umi_counts
)

# Define custom breaks and labels for the axes
x_breaks <- c(10, 100, 1000, 10^4, 10^5, 10^6)
x_labels <- c("10", "100", "1k", "10k", "100k", "1M")

y_breaks <- c(10, 100, 1000, 10^4, 10^5)
y_labels <- c("10", "100", "1k", "10k", "100k")

# Generate the barcode rank plot
ggplot(barcode_data, aes(x = Barcode, y = UMI_Counts)) +
  geom_line() +
  scale_x_log10(breaks = x_breaks, labels = x_labels) +
  scale_y_log10(breaks = y_breaks, labels = y_labels) +
  labs(title = "Barcode Rank Plot",
       x = "Barcodes",
       y = "UMI Counts") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) # Center the plot title

```
```{r}
# save this plot as a png file in ../figures folder with the name Barcode_Rank_Plot.png
ggsave("../figures/Barcode_Rank_Plot.png", width = 10, height = 6, dpi = 300)

```

# EmptyDrops Analysis
## Initial UMI Counts Analysis
The `EmptyDrops` function is used to identify genuine cells from empty droplets based on the distribution of UMI counts across barcodes. The barcodes that correspond to cells are expected to have a higher number of UMIs.

## Step 1: Applying EmptyDrops
We will apply the `EmptyDrops` function to the count data. This function will return a list of likely cell-containing barcodes based on the distribution of UMI counts.

```{r}
# Applying EmptyDrops
# It is assumed that ccRCC_p3 is a Seurat object and the count data are stored accordingly.
# Extract the count matrix from the Seurat object
count_matrix <- GetAssayData(ccRCC_p3, assay = "RNA", layer = "counts")

# Run emptyDrops
emptydrops_results <- emptyDrops(count_matrix)

## show the results again but with decreasing FDR
emptydrops_results[order(emptydrops_results$Total, decreasing = TRUE), ]
```

## Step 2: Filtering Barcodes

Next, we filter the barcodes based on the FDR to retain likely cell-containing droplets. Typically, an FDR threshold of 0.01 is used, but this can be adjusted based on the dataset.

```{r}
# Set the FDR threshold
fdr_threshold <- 0.001

is.cell <- emptydrops_results$FDR <=fdr_threshold
sum(is.cell, na.rm=TRUE)

# Subsetting the matrix to the cell-containing droplets.
# (using 'which()' to handle NAs smoothly).
cell.counts <- count_matrix[,which(is.cell),drop=FALSE]
dim(cell.counts)

```

```{r}

# Check if p-values are lower-bounded by 'niters'
# (increase 'niters' if any Limited==TRUE and Sig==FALSE)
table(Sig=is.cell, Limited=emptydrops_results$Limited)

```

## Step 3: Visualizing the Results After Applying EmptyDrops
We can now visualize the UMI counts of the retained barcodes to compare against the initial UMI counts.
```{r}

# Get the UMI counts for the retained barcodes
retained_barcodes <- which(is.cell)
retained_umi_counts <- total_umi_counts[retained_barcodes]

# Sort the retained UMI counts for better visualization
sorted_retained_umi_counts <- sort(retained_umi_counts, decreasing = TRUE)

```

```{r}

# Create a data frame for the filtered UMI counts
barcode_data_filtered <- data.frame(
  Barcode = seq_along(sorted_retained_umi_counts),
  UMI_Counts = sorted_retained_umi_counts)
```

```{r}

# Define custom breaks and labels for the axes
x_breaks <- c(10, 100, 1000, 10^4, 10^5, 10^6)
x_labels <- c("10", "100", "1k", "10k", "100k", "1M")

y_breaks <- c(10, 100, 1000, 10^4, 10^5)
y_labels <- c("10", "100", "1k", "10k", "100k")

# Generate the barcode rank plot for the filtered data
ggplot(barcode_data_filtered, aes(x = Barcode, y = UMI_Counts)) +
  geom_line() +
    scale_x_log10(breaks = x_breaks, labels = x_labels) +
  scale_y_log10(breaks = y_breaks, labels = y_labels) +
  labs(title = "Barcode Rank Plot After Applying EmptyDrops",
       x = "Barcodes",
       y = "UMI Counts") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))



```

```{r}

initial_umi_plot <- ggplot(barcode_data, aes(x = Barcode, y = UMI_Counts)) +
    geom_line() +
    scale_x_log10(labels = trans_format("log10", math_format(10^.x))) +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    labs(title = "Initial Barcode Rank Plot",
         x = "Barcodes",
         y = "UMI Counts") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.2))

```

```{r}

filtered_umi_plot <- ggplot(barcode_data_filtered, aes(x = Barcode, y = UMI_Counts)) +
    geom_line() +
    scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
    scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
    labs(title = "Barcode Rank After EmptyDrops",
         x = "Barcodes",
         y = "UMI Counts") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.2))


```

```{r plot, out.width='700px', out.height='400px'}
# Assuming you have a plot object for the initial UMI counts named `initial_umi_plot`
# and the plot after filtering named `filtered_umi_plot`

# Using gridExtra
library(gridExtra)

# Create the grid plot
grid_plot <- grid.arrange(initial_umi_plot, filtered_umi_plot, ncol = 2, widths = c(1, 1))

# Set the file path where you want to save the PNG file
file_path <- "../figures/Barcode_Rank_Plot_before_after_EmptyDrops.png"

# Save the grid plot as a PNG file
ggsave(filename = file_path, plot = grid_plot, width = 7, height = 4, units = "in")
 


```



